---
title: "mig package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mig package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: mig.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align="center", 
  fig.width = 5, 
  fig.height = 5
)
```

```{r}
#| label: setup
#| echo: false
#| eval: true
library(mig)
par(bty = "l", pch = 20, xaxs = "i", yaxs = "i")
set.seed(0)
```

The `mig` package provides utilities for kernel density estimation for random vectors  using the multivariate inverse Gaussian distribution defined over the half space $\mathcal{H}_d(\boldsymbol{\beta}) = \{\boldsymbol{x} \in \mathbb{R}^d: \boldsymbol{\beta}^\top \boldsymbol{x} > 0\}$ with location vector $\boldsymbol{\xi}$, scale matrix $\boldsymbol{\Omega}$, whose density is 
\begin{align*}
\frac{\boldsymbol{\beta}^\top\boldsymbol{\xi}}{(2\pi)^{d/2}|\boldsymbol{\Omega}|}(\boldsymbol{\beta}^\top\boldsymbol{x})^{-d/2-1}\exp \left\{-\frac{(\boldsymbol{x} - \boldsymbol{\xi})^\top \boldsymbol{\Omega}^{-1}(\boldsymbol{x}-\boldsymbol{\xi})}{2\boldsymbol{\beta}^\top\boldsymbol{x}}\right\}, \qquad \boldsymbol{x} \in \mathcal{H}_d(\boldsymbol{\beta}).
\end{align*}



## Random number generation
 
 
@Minami:2003 provides a constructive characterization of the inverse Gaussian as the hitting time of a particular hyperplane by a correlated Brownian motion, simulation requires discretization of the latter, and more accurate simulations come at increased costs. 
 
```{r}
#| eval: false
#| echo: false
#| label: fig-penultshape
#| fig-cap: "Penultimate shape behaviour for the inverse Gaussian margin with $\\xi=1$ and \\Omega=1$."
library(statmod)
penult <- mev::smith.penult("invgauss", method = "pot",
                  qu = seq(0.9, 0.9995, by = 0.0005), 
               mean = 1, shape = 1 / 1)
plot(penult$qu, 
     penult$shape, 
     xlab = "quantile level",
     xlim = c(min(penult$qu), 1), 
     ylim = c(0, 0.25),
     yaxs = "i",
     xaxs = "i",
     ylab = "penultimate shape",
     type = 'l',
     bty = "l")
```


Let $\boldsymbol{\beta} \in \mathbb{R}^d$ be the vector defining the halfspace and consider a $(d-1) \times d$ matrix $\mathbf{Q}_2$, such that $\mathbf{Q}_2^\top\boldsymbol{\beta} = \boldsymbol{0}_{d-1}$ and $\mathbf{Q}_2\mathbf{Q}_2^\top = \mathbf{I}_{d-1}$. Theorem 1 (3) of @Minami:2003 implies that, for $\mathbf{Q} = (\boldsymbol{\beta}, \mathbf{Q}_2^\top)\vphantom{Q}^{\top}$ and
\begin{align*} 
Z_1 &\sim \mathsf{MIG}(\boldsymbol{\beta}^\top\boldsymbol{\xi}, \boldsymbol{\beta}^\top\boldsymbol{\Omega}\boldsymbol{\beta}) \\
\boldsymbol{Z}_2 \mid Z_1 = z_1 &\sim \mathsf{Norm}_{d-1}\left[\mathbf{Q}_2\{\boldsymbol{\xi} + \boldsymbol{\Omega}\boldsymbol{\beta}/(\boldsymbol{\beta}^\top\boldsymbol{\Omega}\boldsymbol{\beta})(z_1-\boldsymbol{\beta}^\top\boldsymbol{\xi})\}, z_1(\mathbf{Q}_2\boldsymbol{\Omega}^{-1}\mathbf{Q}_2^\top)^{-1}\right],
\end{align*}
we have $\mathbf{Q}^{-1}\boldsymbol{Z} \sim \mathsf{MIG}(\boldsymbol{\beta}, \boldsymbol{\xi}, \boldsymbol{\Omega})$.

Consider the symmetric orthogonal projection matrix $\mathbf{M}_{\boldsymbol{\beta}}=\mathbf{I}_d - \boldsymbol{\beta}\boldsymbol{\beta}^\top/(\boldsymbol{\beta}^\top\boldsymbol{\beta})$ of rank $d-1$ due to the linear dependency. We build $\mathbf{Q}_2$ from the set of $d-1$ eigenvectors associated to the non-zero eigenvalues of $\mathbf{M}_{\boldsymbol{\beta}}$. We can then perform forward sampling of $Z_1$ and $\boldsymbol{Z}_2 \mid Z_1$ and compute the resulting vectors.

```{r}
#| eval: true
#| echo: true
# Create projection matrix onto the orthogonal complement of beta
d <- 5L # dimension of vector
n <- 1e4L # number of simulations
beta <- rexp(d)
xi <- rexp(d)
Omega <- matrix(0.5, d, d) + diag(d)
# Project onto orthogonal complement of vector
Mbeta <- (diag(d) - tcrossprod(beta)/(sum(beta^2)))
# Matrix is rank-deficient: compute eigendecomposition 
# Shed matrix to remove the eigenvector corresponding to the 0 eigenvalue
Q2 <- t(eigen(Mbeta, symmetric = TRUE)$vectors[,-d])
# Check Q2 satisfies the conditions
all.equal(rep(0, d-1), c(Q2 %*% beta)) # check orthogonality
all.equal(diag(d-1), tcrossprod(Q2)) # check basis is orthonormal
Qmat <- rbind(beta, Q2)
covmat <- solve(Q2 %*% solve(Omega) %*% t(Q2))

# Compute mean and variance for Z1
mu <- sum(beta*xi)
omega <- sum(beta * c(Omega %*% beta))
Z1 <- rmig(n = n, xi = mu, Omega = omega) # uses statmod, with mean = mu and shape mu^2/omega
# Generate Gaussian vectors in two-steps (vectorized operations)
Z2 <- sweep(TruncatedNormal::rtmvnorm(n = n, mu = rep(0, d-1), sigma = covmat), 1, sqrt(Z1), "*")
Z2 <- sweep(Z2, 2, c(Q2 %*% xi), "+") + tcrossprod(Z1 - mu, c(Q2 %*% c(Omega %*% beta)/omega))
# Compute inverse of Q matrix (it is actually invertible)
samp <- t(solve(Qmat) %*% t(cbind(Z1, Z2)))
# Check properties
mle <- mig::fit_mig(samp, beta = beta)
max(abs(mle$xi - xi))
norm(mle$Omega - Omega, type = "f")
max(abs(1 - mle$Omega/Omega))
```

## References
